Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-05-23 paper-new/concrete-model.tex
several fixes to paper

% TODO: later of a predomain
Given a predomain $A$, we can form the predomain $\later A$.
The underlying set is $\later |A|$ and the relation is defined in the obvious way,
i.e., $\tilde{x} \le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
Likewise for bisimilarity.
%
We also give a predomain structure to the natural numbers $\mathbb{N}$, where both the
ordering and the bisimiarity relation are equality.
%
Morphisms of predomains are functions between the underlying sets that preserve the ordering
and the bisimilarity relation. More formally:
%
\begin{definition}
Let $A$ and $A'$ be predomains.
A morphism $f : A \to A'$ is a function between the underlying sets such that for all $x, x'$,
if $x \le_A x'$, then $f(x) \le f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'} f(x')$.
\end{definition}
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error" element $\mho_B \in B$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
\end{itemize}
\end{definition}
%
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B \circ \nxt$.
%
Morphisms of error domains are morphisms of the underlying predomains that preserve the
algebraic structure. More formally:
%
\begin{definition}
Let $B$ and $B'$ be error domains.
A morphism $\phi : B \wand B'$ is a morphism between the underlying predomains such that:
\begin{enumerate}
    \item $\phi(\mho_B) = \mho_{B'}$
    \item $\phi(\theta_B(\tilde{x})) = \theta_{B'}(\lambda t. \phi(\tilde{x}_t))$
\end{enumerate}
\end{definition}
%
We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on the
underlying sets that is downward-closed under $\le_A$ and upward-closed under $\le_{A'}$.
More formally:
%
\begin{definition}
Let $A$ and $A'$ be predomains. A \emph{predomain relation} between $A$ and $A'$
is a relation $R$ between the underlying sets such that:
\begin{enumerate}
    \item (Downward closure): For all $x_1, x_2 \in A$ and $y \in A'$,
    if $x_1 \le_A x_2$ and $x_2 \mathbin{R} y$, then $x_1 \mathbin{R} y$.
    \item (Upward closure): For all $x \in A$ and $y_1, y_2 \in A'$,
    if $x \mathbin{R} y_1$ and $y_1 \le_{A'} y_2$, then $x \mathbin{R} y_2$.
\end{enumerate}
\end{definition}
%
Composition of relations on predomains is the usual relational composition.
Similarly, we define a (monotone) relation on error domains to be a relation on the
underlying predomains that respects error and preserves $\theta$.
%
\begin{definition}
    Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
    $B$ and $B'$ is a relation $R$ between the underlying predomains such that
    \begin{enumerate}
       \item (Respects error): For all $y \in B'$, we have $\mho_B \mathbin{R} y$.
       \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
       if $\later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t )$ then
       $\theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y})$.
    \end{enumerate}
\end{definition}
%
We define composition of error domain relations $R$ on $B_1$ and $B_2$ and $S$
on $B_2$ and $B_3$ to be the least relation containing $R$ and $S$ that respects
error and preserves $\theta$.
Specifically, it is defined inductively by the following rules:
%
\begin{mathpar}
    \inferrule*[right = Comp]
    {b_1 \mathbin{R} b_2 \and b_2 \mathbin{S} b_3}
    {b_1 \mathbin{R \relcomp S} b_3}

@@

This captures the intuition that a program may either return a value,
fail at run-time, or take one or more observable steps of computation.
Previous work has studied such a similar construct, called the guarded lift
monad \cite{mogelberg-paviotti2016}; our version here our version augments it with the notion of error.
Since we claimed that $\li A$ is a monad, we need to define the monadic operations
and show that they respect the monadic laws. The return is just $\eta$, and the monadic extend
is defined via guarded recursion by cases on the input.
Verifying that the monadic laws hold uses \lob-induction and is straightforward.

@@

Explicitly, the ordering on $D$ is given by:
%
\begin{align*}
    \tnat(n) \le \tnat(n') 
        &\iff n = n' \\
    \ttimes (d_1, d_2) \le \ttimes (d_1', d_2')
        &\iff d_1 \le d_2 \text{ and } d_1' \le d_2'\\
    \tfun(\tilde{f}) \le \tfun(\tilde{f'}) 
        &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t)
\end{align*}
%
We define a relation $\inat : \mathbb{N} \rel D$ by
$(n, d) \in \inat$ iff $e_\mathbb{N} \le_D d$.
We similarly define $\itimes : D \times D \rel D$ by
$((d_1, d_2), d) \in \itimes$ iff $e_\times(d_1, d_2) \le_D d$,
and we define $\text{inj}_\to : U(D \arr F D) \rel D$ by
$(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.

>>>>

% TODO: later of a predomain
Given a predomain $A$, we can form the predomain $\later A$.
The underlying set is $\later |A|$ and the relation is defined in the obvious way,
i.e., $\tilde{x} \le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
Likewise for bisimilarity.
%
We also give a predomain structure to the natural numbers $\mathbb{N}$, where both the
ordering and the bisimilarity relation are equality.
%
Morphisms of predomains are functions between the underlying sets that preserve the ordering
and the bisimilarity relation. More formally:
%
\begin{definition}
Let $A$ and $A'$ be predomains.
A morphism $f : A \to A'$ is a function between the underlying sets such that for all $x, x'$,
if $x \le_A x'$, then $f(x) \le f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'} f(x')$.
\end{definition}
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error" element $\mho_B \in B$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
\end{itemize}
\end{definition}
%
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B \circ \nxt$.
%
Morphisms of error domains are morphisms of the underlying predomains that preserve the
algebraic structure. More formally:
%
\begin{definition}
Let $B$ and $B'$ be error domains.
A morphism $\phi : B \wand B'$ is a morphism between the underlying predomains such that:
\begin{enumerate}
    \item $\phi(\mho_B) = \mho_{B'}$
    \item $\phi(\theta_B(\tilde{x})) = \theta_{B'}(\lambda t. \phi(\tilde{x}_t))$
\end{enumerate}
\end{definition}
%
We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on the
underlying sets that is downward-closed under $\le_A$ and upward-closed under $\le_{A'}$.
More formally:
%
\begin{definition}
Let $A$ and $A'$ be predomains. A \emph{predomain relation} between $A$ and $A'$
is a relation $R$ between the underlying sets such that:
\begin{enumerate}
    \item (Downward closure): For all $x_1, x_2 \in A$ and $y \in A'$,
    if $x_1 \le_A x_2$ and $x_2 \mathbin{R} y$, then $x_1 \mathbin{R} y$.
    \item (Upward closure): For all $x \in A$ and $y_1, y_2 \in A'$,
    if $x \mathbin{R} y_1$ and $y_1 \le_{A'} y_2$, then $x \mathbin{R} y_2$.
\end{enumerate}
\end{definition}
%
Composition of relations on predomains is the usual relational composition.
Similarly, we define a (monotone) relation on error domains to be a relation on the
underlying predomains that respects error and preserves $\theta$.
%
\begin{definition}
    Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
    $B$ and $B'$ is a relation $R$ between the underlying predomains such that
    \begin{enumerate}
       \item (Respects error): For all $y \in B'$, we have $\mho_B \mathbin{R} y$.
       \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
       if $\later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t )$ then
       $\theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y})$.
    \end{enumerate}
\end{definition}
%
We define composition of error domain relations $R$ on $B_1$ and $B_2$ and $S$
on $B_2$ and $B_3$ to be the least relation containing $R$ and $S$ that respects
error and preserves $\theta$.
Specifically, it is defined inductively by the following rules:
%
\begin{mathpar}
    \inferrule*[right = Comp]
    {b_1 \mathbin{R} b_2 \and b_2 \mathbin{S} b_3}
    {b_1 \mathbin{R \relcomp S} b_3}

@@

This captures the intuition that a program may either return a value,
fail at run-time, or take one or more observable steps of computation.
Previous work has studied such a similar construct, called the guarded lift
monad \cite{mogelberg-paviotti2016}; our version here augments it with the notion of error.
Since we claimed that $\li A$ is a monad, we need to define the monadic operations
and show that they respect the monadic laws. The return is just $\eta$, and the monadic extend
is defined via guarded recursion by cases on the input.
Verifying that the monadic laws hold uses \lob-induction and is straightforward.

@@

Explicitly, the ordering on $D$ is given by:
%
\begin{align*}
    \tnat(n) \le \tnat(n') 
        &\iff n = n' \\
    \ttimes (d_1, d_2) \le \ttimes (d_1', d_2')
        &\iff d_1 \le d_2 \text{ and } d_1' \le d_2'\\
    \tfun(\tilde{f}) \le \tfun(\tilde{f'}) 
        &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t)
\end{align*}
%
We define a relation $\inat : \mathbb{N} \rel D$ by
$(n, d) \in \inat$ iff $e_\mathbb{N}(n) \le_D d$.
We similarly define $\itimes : D \times D \rel D$ by
$((d_1, d_2), d) \in \itimes$ iff $e_\times(d_1, d_2) \le_D d$,
and we define $\text{inj}_\to : U(D \arr F D) \rel D$ by
$(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.

### 2024-05-23 paper-new/extensional-model.tex
several fixes to paper

In order to organize our construction of denotational models we first
develop sufficient \emph{abstract} categorical semantics of gradually
typed languages. We start by modeling the type and term structure of
gradual typing and then extend this to type and term precision.
%
Gradually typed languages inherently involve computational effects of
errors and non-termination and typically in practice many others such
as mutable state and I/O.
%
To model this cleanly categorically, we follow New, Licata and Ahmed's
GTT calculus and base our models off of Levy's Call-by-push-value
(CBPV) calculus which is a standard model of effectful programming
\cite{levy99}.
%
There are several notions of model of CBPV from the literature with
varying requirements of which connectives are present
\cite{levy99,cfm2016,eec}, we will use a variant which models precisely
the connectives we require and no more
($1,\times,F,U,\to$)\footnote{It is essential in this case that we do
not require a cartesian closed category of values as there is no way
to implement casts for an exponential in general.}.

@@

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras of the monad, defining $A
  \to B$ as the powering of algebras, $F$ as the free algebra and $U$
  as the underlying object functor.
\end{example}

To additionally model the error terms, we add a requirement that there
is a natural transformation $\mho : 1 \to U B$. The naturality
requirement encodes that strict morphisms (e.g., the denotations of
evaluation contexts) preserve errors.

We can then model CBV terms and types in a straightforward adaptation
of Levy's interpretation of CBV in CBPV. We interpret types $A$ and
contexts $\Gamma$ as objects $A,\Gamma \in \mathcal V$ and CBV terms
$\Gamma \vdash M : A$ as morphisms of any of the equivalent forms
$\mathcal E(F\Gamma, FA) \cong \mathcal V(\Gamma \vdash UFA) \cong
\mathcal E(F(1), \Gamma \to F(A))$. The most interesting type
translation is the CBV function type: $A \ra A' = U(A \to F A')$.
%
Such a model validates all type-based equational reasoning, i.e.,
$\beta\eta$ equality, and models the introduction and elimination
rules for CBV.
%
Thus a CBPV model is sufficient to interpret the CBV term language. We
will require additional structure to interpret the precision and type
casts.

@@

New and Licata modeled the graduality and type casts for call-by-name
gradual typing using \emph{double categories}, which are defined to be
categories internal to the category of categories. That is, a double
category $\mathcal C$ consists of a category $\mathcal C_o$ of
``objects and function morphisms'' and a category $\mathcal C_{sq}$ of
``relation morphisms and squares'' with functors (reflexive relation)
$r : C_o \to C_{sq}$ and (source and target) $s,t : C_{sq} \to C_o$
satisfying $sr = tr = \id$ as well as a composition operation $c :
C_{sq} \times_{s,t} C_{sq} \to C_{sq}$ respecting source and
target. This models an abstract notion of functions and relations. For
notation, we write function morphisms as $f : A \to B$ and relation
morphisms as $c : A \rel B$ where $c \in C_{sq}$ and $s(c) = A$ and
$t(c) = B$. Finally a morphism $\alpha$ from $c$ to $d$ with
$s(\alpha) = f$ and $s(\beta) = g$ is visualized as
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBIl0sWzEsMCwiQiJdLFswLDEsIkEnIl0sWzEsMSwiQiciXSxbMCwyLCJmIiwyXSxbMSwzLCJnIl0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	A \& B \\
	{A'} \& {B'}
	\arrow["f"', from=1-1, to=2-1]
	\arrow["g", from=1-2, to=2-2]
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
And is thought of as an abstraction of the notion of relatedness of
functions: functions take related inputs to related outputs. The
composition operations and functoriality give us a notion of
composition of relations as well as functions and vertical and
horizontal composition of squares. In this work we will be chiefly
interested in \emph{locally thin} double categories, that is, double
categories where there is at most one square for any $f,c,g,d$. In
this case we use the notation $f \leq_{c,d} g$ to mean that a square
like the above exists.

New, Licata and Ahmed \cite{new-licata-ahmed2019} extended the axiomatic
syntax to call-by-push-value but did not analyze the structure
categorically. We fill in this missing analysis now: a model of the
congruence rules of their system can be given by a locally thin
``double CBPV model'', which we define as a category internal to the
category of CBPV models and \emph{strict} homomorphisms of CBPV
models\footnote{it may be possible to also define this as a notion of
CBPV model internal to some structured $2$-category of categories, but
the authors are not aware of any such definition of an internal CBPV
model}. A strict homomorphism of CBPV models from $(\mathcal
V,\mathcal E,\ldots)$ to $(\mathcal V', \mathcal E',\ldots)$ consists
of functors $G_v : \mathcal V \to \mathcal V'$ and $G_e : \mathcal E
\to \mathcal E'$ that strictly preserve all CBPV constructions, see
the appendix for a more detailed definition. We call this a strict morphism in contrast to a \emph{lax} morphism, which only preserves CBPV constructions up to transformation.
Some of the data of a double CBPV model can be visualized as follows:
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHZzcSJdLFsyLDAsIlxcZXNxIl0sWzAsMiwiXFx2ZiJdLFsyLDIsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFswLDIsIlxcdHYiLDIseyJjdXJ2ZSI6Mn1dLFsxLDMsIlxcc2UiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHRlIiwyLHsiY3VydmUiOjJ9XSxbMywxLCJcXHJlIl0sWzQsNSwiXFxib3QiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	\vsq \&\& \esq \\
	\\
	\vf \&\& \ef
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=3-1, to=3-3]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=3-3, to=3-1]
	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=1-1, to=1-3]
	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=1-3, to=1-1]
	\arrow["\rv", from=3-1, to=1-1]
	\arrow["\sv", curve={height=-12pt}, from=1-1, to=3-1]
	\arrow["\tv"', curve={height=12pt}, from=1-1, to=3-1]
	\arrow["\se", curve={height=-12pt}, from=1-3, to=3-3]
	\arrow["\te"', curve={height=12pt}, from=1-3, to=3-3]
	\arrow["\re", from=3-3, to=1-3]
	\arrow["\bot"{description}, draw=none, from=0, to=1]
	\arrow["\bot"{description}, draw=none, from=2, to=3]
\end{tikzcd}\]
Type precision $A \ltdyn A'$ is interpreted as a relation morphism
$c_A : A \rel A'$ in $\mathcal V_{sq}$, and term precision $\Gamma
\ltdyn \Gamma' \vdash M \ltdyn M' : A \ltdyn A'$ is interpreted as a
square $M \ltdyn_{c_\Gamma,UF c_A} M'$. The fact that $t,r$ and the
composition are all given by strict CBPV homomorphisms says that all
the type constructors lift to precision (monotonicity of type
constructors) as well as all term constructors (congruence). Further,
$r$ and composition being strict homomorphisms implies that all type
constructors strictly preserve the identity relation (identity
extension) and composition.

@@

  Dually, $c : A \rel B$ is right representable by $g : B \to A$ if
  $\id \ltsq{c}{r(A)} g$ and $g \ltsq{r(B)}{c} \id$.
\end{definition}
These rules are sufficient to model the UpL/UpR/DnL/DnR rules for
casts. Additionally, since representable morphisms compose and so the
compositionality of casts comes for free. However, the retraction
property must be added as an additional axiom to the model.  To model
the error being a least element we add the requirement that $\mho
\circ ! \ltsq{r(A)}{r(UB)} f$ holds for all $f : \mathcal V(A,B)$.
Finally, the dynamic type can be modeled as an arbitrary value type $D$ with
arbitrary relations $\nat \rel D$ and $D \ra D \rel D$ and $D \times D
\rel D$ (or whatever basic type cases are required).

>>>>

In order to organize our construction of denotational models, we first
develop sufficient \emph{abstract} categorical semantics of gradually
typed languages. We start by modeling the type and term structure of
gradual typing and then extend this to type and term precision.
%
Gradually typed languages inherently involve computational effects of
errors and non-termination and typically in practice many others such
as mutable state and I/O.
%
To model this cleanly categorically, we follow New, Licata and Ahmed's
GTT calculus and base our models on Levy's Call-by-push-value
(CBPV) calculus which is a standard model of effectful programming
\cite{levy99}.
%
There are several notions of model of CBPV from the literature with
varying requirements of which connectives are present
\cite{levy99,cfm2016,eec}. We will use a variant which models precisely
the connectives we require and no more
($1,\times,F,U,\to$)\footnote{It is essential in this case that we do
not require a cartesian closed category of values as there is no way
to implement casts for an exponential in general.}.

@@

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras of the monad, defining $A
  \to B$ as the powering of algebras, $F$ as the free algebra functor and $U$
  as the underlying object functor.
\end{example}

To additionally model the error terms, we add a requirement that there
is a natural transformation $\mho : 1 \Rightarrow U$. The naturality
requirement encodes that strict morphisms (e.g., the denotations of
evaluation contexts) preserve errors.

We can then model CBV terms and types in a straightforward adaptation
of Levy's interpretation of CBV in CBPV. We interpret types $A$ and
contexts $\Gamma$ as objects $A,\Gamma \in \mathcal V$ and CBV terms
$\Gamma \vdash M : A$ as morphisms of any of the equivalent forms
$\mathcal E(F\Gamma, FA) \cong \mathcal V(\Gamma , UFA) \cong
\mathcal E(F(1), \Gamma \to F(A))$. The most interesting type
translation is the CBV function type: $A \ra A' = U(A \to F A')$.
%
Such a model validates all type-based equational reasoning, i.e.,
$\beta\eta$ equality, and models the introduction and elimination
rules for CBV.
%
Thus a CBPV model is sufficient to interpret the CBV term language. We
will require additional structure to interpret the precision and type
casts.

@@

New and Licata \cite{new-licata18} modeled the graduality and type casts for call-by-name
gradual typing using \emph{double categories}, which are defined to be
categories internal to the category of categories. That is, a double
category $\mathcal C$ consists of a category $\mathcal C_o$ of
``objects and function morphisms'' and a category $\mathcal C_{sq}$ of
``relation morphisms and squares'' with functors (reflexive relation)
$r : C_o \to C_{sq}$ and (source and target) $s,t : C_{sq} \to C_o$
satisfying $sr = tr = \id$ as well as a composition operation $c :
C_{sq} \times_{s,t} C_{sq} \to C_{sq}$ respecting source and
target. This models an abstract notion of functions and relations. For
notation, we write function morphisms as $f : A \to B$ and relation
morphisms as $c : A \rel B$ where $c \in C_{sq}$ and $s(c) = A$ and
$t(c) = B$. Finally, a morphism $\alpha$ from $c$ to $d$ with
$s(\alpha) = f$ and $s(\beta) = g$ is visualized as
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBIl0sWzEsMCwiQiJdLFswLDEsIkEnIl0sWzEsMSwiQiciXSxbMCwyLCJmIiwyXSxbMSwzLCJnIl0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	A \& B \\
	{A'} \& {B'}
	\arrow["f"', from=1-1, to=2-1]
	\arrow["g", from=1-2, to=2-2]
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
Such a morphism is thought of as an abstraction of the notion of relatedness of
functions: functions take related inputs to related outputs. The
composition operations and functoriality give us a notion of
composition of relations as well as functions and vertical and
horizontal composition of squares. In this work we will be chiefly
interested in \emph{locally thin} double categories, that is, double
categories where there is at most one square for any $f,c,g,d$. In
this case we use the notation $f \leq_{c,d} g$ to mean that a square
like the above exists.

New, Licata and Ahmed \cite{new-licata-ahmed2019} extended the axiomatic
syntax to call-by-push-value but did not analyze the structure
categorically.
We fill in this missing analysis now, extending the double categorical semantics
given in \cite{new-licata18} from cartesian closed categories to CBPV models.
A model of the congruence rules of the CBPV system can be given by a locally thin
``double CBPV model'', which we define as a category internal to the
category of CBPV models and \emph{strict} homomorphisms of CBPV
models\footnote{it may be possible to also define this as a notion of
CBPV model internal to some structured $2$-category of categories, but
the authors are not aware of any such definition of an internal CBPV
model}. A strict homomorphism of CBPV models from $(\mathcal
V,\mathcal E,\ldots)$ to $(\mathcal V', \mathcal E',\ldots)$ consists
of functors $G_v : \mathcal V \to \mathcal V'$ and $G_e : \mathcal E
\to \mathcal E'$ that strictly preserve all CBPV constructions, see
the appendix for a more detailed definition. We call this a strict morphism
in contrast to a \emph{lax} morphism, which only preserves CBPV constructions
up to transformation.
Some of the data of a double CBPV model can be visualized as follows:
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHZzcSJdLFsyLDAsIlxcZXNxIl0sWzAsMiwiXFx2ZiJdLFsyLDIsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFswLDIsIlxcdHYiLDIseyJjdXJ2ZSI6Mn1dLFsxLDMsIlxcc2UiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHRlIiwyLHsiY3VydmUiOjJ9XSxbMywxLCJcXHJlIl0sWzQsNSwiXFxib3QiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	\vsq \&\& \esq \\
	\\
	\vf \&\& \ef
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=3-1, to=3-3]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=3-3, to=3-1]
	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=1-1, to=1-3]
	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=1-3, to=1-1]
	\arrow["\rv", from=3-1, to=1-1]
	\arrow["\sv", curve={height=-12pt}, from=1-1, to=3-1]
	\arrow["\tv"', curve={height=12pt}, from=1-1, to=3-1]
	\arrow["\se", curve={height=-12pt}, from=1-3, to=3-3]
	\arrow["\te"', curve={height=12pt}, from=1-3, to=3-3]
	\arrow["\re", from=3-3, to=1-3]
	\arrow["\bot"{description}, draw=none, from=0, to=1]
	\arrow["\bot"{description}, draw=none, from=2, to=3]
\end{tikzcd}\]
Type precision $A \ltdyn A'$ is interpreted as a relation morphism
$c_A : A \rel A'$ in $\mathcal V_{sq}$, and term precision $\Gamma
\ltdyn \Gamma' \vdash M \ltdyn M' : A \ltdyn A'$ is interpreted as a
square $M \ltdyn_{c_\Gamma,UF c_A} M'$. The fact that $t,r$ and the
composition are all given by strict CBPV homomorphisms says that all
the type constructors lift to precision (monotonicity of type
constructors) as well as all term constructors (congruence). Further,
$r$ and composition being strict homomorphisms implies that all type
constructors strictly preserve the identity relation (identity
extension) and composition.

@@

  Dually, $c : A \rel B$ is right representable by $g : B \to A$ if
  $\id \ltsq{c}{r(A)} g$ and $g \ltsq{r(B)}{c} \id$.
\end{definition}
These rules are sufficient to model the UpL/UpR/DnL/DnR rules for
casts. Additionally, since representable morphisms compose and so the
compositionality of casts comes for free. However, the retraction
property must be added as an additional axiom to the model.  To model
the error being a least element we add the requirement that $\mho
\circ {!} \ltsq{r(A)}{r(UB)} f$ holds for all $f : \mathcal V(A,B)$.
Finally, the dynamic type can be modeled as an arbitrary value type $D$ with
arbitrary relations $\nat \rel D$ and $D \ra D \rel D$ and $D \times D
\rel D$ (or whatever basic type cases are required).

### 2024-05-23 paper-new/intro.tex
several fixes to paper

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
In modern programming language design, there is a tension between \emph{static} typing
and \emph{dynamic} typing disciplines.
With static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static 
typing, the programmer is assured that if the program passes the type-checker, their
program is free of type errors, and moreover, soundness of the equational theory implies
that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their
functions. Most languages choose between static or dynamic typing and as a result, programmers that initially write their code in a dynamically typed language
in order to benefit from faster prototyping and development time need to rewrite the
some or all of their codebase in a static language if they would like to receive the benefits of static
typing once their codebase has matured.

@@

In this paper, we develop an adequate denotational semantics that
satisfies graduality and soundness of the equational theory of cast
calculi using the techniques of SGDT.  Our longer-term goal is to
mechanize these proofs in a reusable way in the Guarded Cubical Agda
proof assistant, thereby providing a framework to use to more easily
and conveniently prove that existing languages satisfy graduality and
have sound equational theories. Moreover, the aim is for designers of
new languages to utilize the framework to facilitate the design of new
provably-correct gradually-typed languages with more complex features.

@@

The main contribution of this work is a categorical and denotational
semantics for gradually typed langauges that models not just the term
language but the graduality property as well.
\begin{enumerate}
\item First, we give a simple abstract categorical model of GTT using CBPV double categories.
\item Next, we modify this semantics to develop reflexive graph- and
  double categorical models that abstract over the details of
  step-indexed models, and provide a method for constructing such models.
\item We instantiate the abstract construction to provide a concrete
  semantics in informal guarded type theory.
\item We prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision
  $M \ltdyn N : \nat$ has the expected semantics, that $M$ errors or
  $M$ and $N$ have the same extensional behavior.
\end{enumerate}

>>>>

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
In modern programming language design, there is a tension between \emph{static} typing
and \emph{dynamic} typing disciplines.
With static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static 
typing, the programmer is assured that if the program passes the type-checker, their
program is free of type errors, and moreover, soundness of the equational theory implies
that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their
functions. Most languages choose between static or dynamic typing and as a result, programmers that initially write their code in a dynamically typed language
in order to benefit from faster prototyping and development time need to rewrite
some or all of their codebase in a static language if they would like to receive the benefits of static
typing once their codebase has matured.

@@

In this paper, we develop an adequate denotational semantics that
satisfies graduality and soundness of the equational theory of cast
calculi using the techniques of SGDT.  Our longer-term goal is to
mechanize these proofs in a reusable way in the Guarded Cubical Agda
proof assistant, thereby providing a framework to more easily
and conveniently prove that existing languages satisfy graduality and
have sound equational theories. Moreover, the aim is for designers of
new languages to utilize the framework to facilitate the design of new
provably-correct gradually-typed languages with more complex features.

@@

The main contribution of this work is a categorical and denotational
semantics for gradually typed langauges that models not just the term
language but the graduality property as well.
\begin{enumerate}
\item First, we give a simple abstract categorical model of Gradual Type Theory using CBPV double categories.
\item Next, we modify this semantics to develop reflexive graph- and
  double categorical models that abstract over the details of
  step-indexed models, and provide a method for constructing such models.
\item We instantiate the abstract construction to provide a concrete
  semantics in informal guarded type theory.
\item We prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision
  $M \ltdyn N : \nat$ has the expected semantics, that $M$ errors or
  $M$ and $N$ have the same extensional behavior.
\end{enumerate}
